SafeNet: A European database of plant-pollinator interactions

### 2024-02-23 Manuscript/Ms.qmd
Intro-->Ready. Polishing methods and jumping to results

**Motivation:** Pollinators play a crucial role in maintaining Earth's terrestrial biodiversity and human food production by mediating sexual reproduction for most flowering plants. However, their diversity and role as pollinators are increasingly compromised by rapid human-induced environmental changes. One of the major challenges for pollinator conservation, is the lack of robust generalisable data across space and time to comprehend the conservation status and population trends among different pollinator species. Here, we present the SafeNet database, a fully reproducible European-level database containing harmonized taxonomic data on plant-pollinator interactions referenced in both space and time. This database offers an open workflow that allows researchers to track decisions and edit them according to their preferences, while also providing other ecological variables of interest. We hope this database can help researchers to: 1) identify taxonomic, ecological and geographical gaps of knowledge on plant-pollinator interactions; and 2) explore the impacts of global change to guide future conservation planning for both plant and pollinator species and their interactions.

@@

**Spatial location and grain:** The database contains `r n_sites` different sampling locations from natural and anthropogenic habitats that fall in `r n_bioregion` different bio-climatic regions. All records are geo-referenced and presented in the World Geodetic System 1984 (WGS84).

@@

Europe is one of the continents with a larger amount of available biodiversity data [@proencca2017], yet still exhibits major gaps [@wetzel2018; @bennett2018]. While species checklists need to be treated carefully, especially at a macro-ecological scale [@grenie2023], the growing number of European plant and pollinator checklists along with occurrence data, is setting a foundation for the conservation of its flora and their pollinators. However, species richness is just one component of biodiversity and documenting the interaction between plants and pollinators is essential for understanding the fate of the species within an ecosystem [@jordano2016]. Numerous works have studied plant-pollinator interactions in the last decades, originating thousands of plant-pollinator interaction networks worldwide. Several initiatives have tried to integrate plant-pollinator interaction data into databases such as *Mangal* [@poisot2016] or *GloBI* [@poelen2014], resulting in numerous large scale comparative analyses that have enhanced our understanding of the ecology of plants and pollinators [e.g., European wild bee data trends\; @marshall2024]. A major drawback of these global databases is the limited number of studies at regional scales and the lack of additional information (e.g., sampling effort or methodology), which hampers our understanding of plant-pollinator interactions. Thus, in a scenario of rapid environmental change, integrating robust informative plant-pollinator interaction data is essential for evaluating our current knowledge at the European level, guiding conservation planning, and establishing a solid foundation for global change research.

3rd paragraph

Low connectance?

Species occurrences

Threatened species? Red lists?

Discuss actual TRENDS!

4paragraph

Here, we present the SafeNet database, an acronym derived from the European project Safeguard and plant-pollinator networks, which contains harmonized informative interaction data of plants and pollinators at European level. The primary focus of the pollinator taxonomic groups of this database are Hymenoptera, Diptera, Lepidoptera and Coleoptera.

@@

@@

This European metaweb consist of datasets of plant-pollinator interactions compiled initially by a wide number of researchers and institutions within the European continent. This dataset consist of `r study_number` independent published and unpublished studies conducted during the time period `r min_year` - `r max_year` on `r country_number` different countries (**Figure 1a**). The dataset contains a total of `r prettyNum(interaction_number, big.mark=",")` distinct interactions, considering interaction as the contact of a given pollinator to the reproductive structure of a particular plant, from `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator and `r prettyNum(plant_spp_number, big.mark=",")` plant species. The majority of plant and pollinator species tend to be regionally specific, with only a minor portion of them being shared across a broad range of studies (**Figure 1b**). These different studies differ in sampling effort and methodology, although most studies took place within a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%), sampled a given location an average of `r format(round(average_sampling_days, 2), nsmall=2)` days, and documented interactions mostly by using `r  most_common_method[[1]]` as sampling method (`r format(round(most_common_method[[3]], 2), nsmall = 2)`%).

Define study area//Data acquisition//Data structure?

TABLE? Describe cols?

>>>>

**Motivation:** Pollinators play a crucial role in maintaining Earth's terrestrial biodiversity and human food production by mediating sexual reproduction for most flowering plants. However, their diversity and role as pollinators are increasingly compromised by rapid human-induced environmental changes. One of the major challenges for pollinator conservation, is the lack of robust generalisable data across space and time to comprehend the conservation status and population trends among different pollinator species. Here, we present the SafeNet database, a fully reproducible European-level database containing harmonized taxonomic data on plant-pollinator interactions referenced in both space and time. This database offers an open workflow that allows researchers to track decisions and edit them according to their preferences, while also providing other ecological variables of interest. We hope this database can help researchers to: 1) identify taxonomic, ecological and geographical gaps of knowledge on plant-pollinator interactions; and 2) explore the impacts of global change to guide future conservation planning for both plant and pollinator species.

@@

**Spatial location and grain:** The database contains `r prettyNum(n_sites, big.mark=",")` different sampling locations from natural and anthropogenic habitats that fall in `r n_bioregion` different bio-climatic regions. All records are geo-referenced and presented in the World Geodetic System 1984 (WGS84).

@@

Europe is one of the continents with a larger amount of available biodiversity data [@proencca2017], yet still exhibits major gaps [@wetzel2018; @bennett2018]. While species checklists need to be treated carefully, especially at a macro-ecological scale [@grenie2023], the growing number of European plant and pollinator checklists along with occurrence data, is setting a foundation for the conservation of its flora and their pollinators. However, species richness is just one component of biodiversity and documenting the interaction between plants and pollinators is essential for understanding the fate of the species within an ecosystem [@jordano2016]. Numerous works have studied plant-pollinator interactions in the last decades, originating thousands of plant-pollinator interaction networks worldwide. Several initiatives have tried to integrate plant-pollinator interaction data into databases such as *Mangal* [@poisot2016] or *GloBI* [@poelen2014], resulting in numerous large scale comparative analyses that enhanced our understanding of the ecology of plants and pollinators [e.g., European wild bee data trends\; @marshall2024]. Despite all these resources, Europe lacks accessible harmonized plant-pollinator interaction data that allow researchers to evaluate the state of the art of plant-pollinator interactions at European level, which will guide research efforts, conservation planning and will set a foundation for future global change research.

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties across them, such as truncated power-law degree distributions [@jordano1987] or nestedness [@bascompte2003], and has added relevant information of their interaction ecology of plant-pollinator networks across geographic regions [@olesen2002; @traveset2016] and environmental gradients [@ramos2010; @rech2016; @saunders2023]. Although macro-ecological approaches make significant contributions to knowledge, they tend to be rare, and are strongly influenced by the spatio-temporal availability and nature of the data [@burkle2011; @trojelsgaard2013]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which can impact the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016]. Further, despite little sampling effort could capture most of the relevant functional species [@hegland2010], most plant-pollinator networks have unobserved interactions because undersampling [@olesen2011; @chacoff2012]. Thus, the lack of strong spatio-temporal coverage along the current intrinsic limitations of sampling plant-pollinator networks, highlight the need of keep gathering and integrating informative species interaction data to properly unravel the different ecological processes at large scales.

Here, we present the SafeNet database, an acronym derived from the European project Safeguard and plant-pollinator networks, which contains harmonized informative interaction data of plants and pollinators at European level. The primary focus of the pollinator taxonomic groups is on the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@potts2015], which comprise almost the totality of recorded interactions in the SafeNet database (`r format(round(main_orders_percent, 2), nsmall = 2)`%). SafeNet aims to cover a wide range of taxonomic groups and habitats while also provides other variables of interest that allow to better control for the ecological context and sampling methods. In addition, SafeNet offers a transparent and accessible workflow of its data management and species harmonization that allow to being reused and keep building on it over time. We expect that this database will help to evaluate macro-ecological processes and current gaps of plant-pollinator interactions at European level.

@@

## Data acquisition

This database is the result of one of the working packages of the European project Safeguard. The SafeNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent. Data acquisition followed a non-systematic approach. First, data was directly asked to members of the Safeguard project and then, the request was extended to data owners outside of the project. These other studies were identified by direct communication with other colleagues and by directly searching studies on Google Scholar of under-represented regions or countries within the database. To keep high quality standards that will allow robust future ecological research, we only included: 1) studies that contained time- and geo-referenced records of plant-pollinator interactions; 2) studies with phyto-centric plant-pollinator networks with quantitative visitation data. The database contains `r study_number` independent published and unpublished studies conducted during the time period `r min_year` - `r max_year` on `r country_number` different countries (**Figure 1a**).

@@

SafeNet includes a total of `r prettyNum(interaction_number, big.mark=",")` distinct interactions, considering interaction as the contact of a given pollinator to the reproductive structure of a particular plant, from `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator and `r prettyNum(plant_spp_number, big.mark=",")` plant species. The majority of plant and pollinator species tend to be regionally specific, with only a minor portion of them being shared across a broad range of studies (**Figure 1b**). These different studies differ in sampling effort and methodology, although most studies took place within a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%), sampled a given location an average of `r format(round(average_sampling_days, 2), nsmall=2)` days, and documented interactions mostly by using `r  most_common_method[[1]]` as sampling method (`r format(round(most_common_method[[3]], 2), nsmall = 2)`%).

Define study area//Data structure? TABLE? Describe cols?
